En Garde! Unguarded Iteration for Reversible Computation in the Delay Monad\thanks{Niccol\`o Veltri was supported by a research grant (13156) from VILLUM FONDEN.}

### 2019-07-01 mpc19/paper.tex
halfway to final version

\Pio{} is a family of reversible combinator calculi comprising structural
isomorphisms and combinators corresponding to those found in dagger-traced
$\omega$-continuous rig categories~\cite{Karvonen19} (a kind of dagger category
with a trace, monoidal sums $\oplus$ and products $\oplus$ such that they form
a rig structure, and fixed points of the functors formed from the rig
structure). Though superficially simple, \Pio{} is expressive enough as a
metalanguage to give semantics to the typed reversible functional programming
language Theseus~\cite{JamesS14}.

@@

Throughout the paper, we reason constructively in Martin-L\"of type
theory. Classically, the delay monad is isomorphic to the maybe monad
$\Maybe\,X \equiv X + 1$,
and thus just a complication of something that can be expressed much
simpler. Constructively, however, they are very different. In particular,
it is impossible to define a well-behaved trace combinator for the
maybe monad without assuming classical principles such as the limited
principle of omniscience.

@@

%% \begin{align*}
%% & \iterD' : (A \to \Delay (A + B)) \to \Delay (A + B) \to \Delay B \\
%% & \iterD' \,f \,(\now \,(\inl\, a)) = \later \,(\iterD' \,f \,(f \,a)) \\
%% & \iterD' \,f \,(\now \,(\inr\, b)) = \now \,b \\
%% & \iterD' \,f \,(\later \,x) = \later \,(\iterD' \,f \,x) \\ 
%% & \\
%% & \iterD : (A \to \Delay (A + B)) \to A \to \Delay B \\
%% & \iterD \,f \,a = \iterD' \,f \,(f \,a)
%% \end{align*}
The definition above can be given the following intuitive
explanation. If $f\,a$ does not terminate, then $\iterD\,f\,a$ does
not terminate as well. If $f\,a$ terminates, there are two
possibilities: either $f\,a$ converges to $\inr\,b$, in which case
$\iterD\,f\,a$ terminates returning the value $b$; or $f\,a$ converges
to $\inl\,a'$, in which case we repeat the procedure by replacing $a$
with $a'$. Notice that in the latter case we also add one occurrence
of $\later$ to the total computation time. This addition is necessary
for ensuring the productivity of the corecursively defined function
$\iterD'$. In fact, by changing the second line of its specification to
$\iterD' \,f \,(\now \,(\inr\, a)) = \iterD' \,f \,(f a)$
and taking $f = \inrD$, we would have that $\iterD' \,f \,(\now
\,(\inr\, a))$ unfolds indefinitely without producing any output. In
Agda, such a definition would be rightfully rejected by the termination
checker.

@@

In this section, we provide some details on the interpretation of the
syntax of \Pio, presented in Section \ref{sec:syntax}, into the
category $\Inv\Dapprox$. Types of \Pio\ are modelled as objects of
$\Inv\Dapprox$, which are types of the metatheory. In Agda, the
interpretation of types $\semTy -$ takes in input a \Pio\ type
$A : \Ty\,n$ and an environment $\rho : \mathsf{Fin}\,n \to \Set$
giving semantics to each variable in context. The interpretation is
mutually inductively defined together with the operation $\semMu -$
giving semantics to the $\mu$ type former.
\[
\begin{array}{lll}
\semTy \Z & \rho & = 0 \\
\semTy {A \oplus B} & \rho & = \semTy A \,\rho+ \semTy B  \, \rho\\
\semTy \I & \rho & = 1 \\
\semTy {A \otimes B} & \rho & = \semTy A \,\rho \times \semTy B \,\rho\\
\semTy {\Var\,i} & \rho & = \rho\, i \\
\semTy {\mu A} & \rho & = \semMu A \,\rho
\end{array}
\qquad \qquad
\infer{\mathsf{semFold} \,x : \semMu A\,\rho}{x : \semTy A \, \rho}
\]

>>>>

\Pio{} is a family of reversible combinator calculi comprising structural
isomorphisms and combinators corresponding to those found in dagger-traced
$\omega$-continuous rig categories~\cite{Karvonen19} (a kind of dagger category
with a trace, monoidal sums $\oplus$ and products $\otimes$ such that they form
a rig structure, and fixed points of the functors formed from the rig
structure). Though superficially simple, \Pio{} is expressive enough as a
metalanguage to give semantics to the typed reversible functional programming
language Theseus~\cite{JamesS14}.

@@

Throughout the paper, we reason constructively in Martin-L\"of type
theory. Classically, the delay monad (quotiented by
weak bisimilarity)
is isomorphic to the maybe monad
$\Maybe\,X \equiv X + 1$,
and thus just a complication of something that can be expressed much
simpler. Constructively, however, they are very different. In particular,
it is impossible to define a well-behaved trace combinator for the
maybe monad without assuming classical principles such as the limited
principle of omniscience.

@@

%% \begin{align*}
%% & \iterD' : (A \to \Delay (A + B)) \to \Delay (A + B) \to \Delay B \\
%% & \iterD' \,f \,(\now \,(\inl\, a)) = \later \,(\iterD' \,f \,(f \,a)) \\
%% & \iterD' \,f \,(\now \,(\inr\, b)) = \now \,b \\
%% & \iterD' \,f \,(\later \,x) = \later \,(\iterD' \,f \,x) \\ 
%% & \\
%% & \iterD : (A \to \Delay (A + B)) \to A \to \Delay B \\
%% & \iterD \,f \,a = \iterD' \,f \,(f \,a)
%% \end{align*}
The definition above can be given the following intuitive
explanation. If $f\,a$ does not terminate, then $\iterD\,f\,a$ does
not terminate as well. If $f\,a$ terminates, there are two
possibilities: either $f\,a$ converges to $\inl\,b$, in which case
$\iterD\,f\,a$ terminates returning the value $b$; or $f\,a$ converges
to $\inr\,a'$, in which case we repeat the procedure by replacing $a$
with $a'$. Notice that in the latter case we also add one occurrence
of $\later$ to the total computation time. This addition is necessary
for ensuring the productivity of the corecursively defined function
$\iterD'$. In fact, by changing the second line of its specification to
$\iterD' \,f \,(\now \,(\inr\, a)) = \iterD' \,f \,(f a)$
and taking $f = \inrD$, we would have that $\iterD' \,f \,(\now
\,(\inr\, a))$ unfolds indefinitely without producing any output. In
Agda, such a definition would be rightfully rejected by the termination
checker.

@@

In this section, we provide some details on the interpretation of the
syntax of \Pio, presented in Section \ref{sec:syntax}, into the
category $\Inv\Dapprox$. Types of \Pio\ are modelled as objects of
$\Inv\Dapprox$, which are types of the metatheory. In Agda, the
interpretation of types $\semTy -$ takes in input a \Pio\ type
$A : \Ty\,n$ and an environment $\rho : \mathsf{Fin}\,n \to \Set$
giving semantics to each variable in context. The interpretation is
mutually inductively defined together with the operation $\semMu -$
giving semantics to the $\mu$ type former. Remember that, given $A :
\Ty\,(n+1)$ and $B : \Ty\,n$, we write $\sub\,A \,B$ for the
substituted type $A[B/X]$, where $X$ corresponds to the $(n+1)$-th
variable in context.

\[
\begin{array}{lll}
\semTy \Z & \rho & = 0 \\
\semTy {A \oplus B} & \rho & = \semTy A \,\rho+ \semTy B  \, \rho\\
\semTy \I & \rho & = 1 \\
\semTy {A \otimes B} & \rho & = \semTy A \,\rho \times \semTy B \,\rho\\
\semTy {\Var\,i} & \rho & = \rho\, i \\
\semTy {\mu A} & \rho & = \semMu A \,\rho
\end{array}
\qquad \qquad
\infer{\mathsf{semFold} \,x : \semMu {A}\,\rho}{x : \semTy
  {\sub\,A\,(\mu A)} \, \rho}
\]
